---
title: "P8105 HW2 - Sophie Gakis (sig2127)"
output: github_document
date: "2026-10-01"
---

# Problem 1 

Loading the tidyverse to start

```{r}
library(tidyverse)
```

## Importing the NYC Transit dataset, reading, and cleaning the data.  

```{r}
transit_df = 
  read_csv(
    file = "Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
    janitor::clean_names() |>
    select(line, station_name, station_latitude, station_longitude,  starts_with("route"), entry, vending, entrance_type, ada) |>
    mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE 
    )
  )

nrow(transit_df)
ncol(transit_df)

view(transit_df)
```

## Writing a paragraph to describe the data 

The transit_df dataset now contains 19 variables (or columns). It contains subway lines, station names, latitude/longitude values for each station, what routes are served by each station, what entrace type is in the station, if there is vending and entry at the station, and if the station is ADA compliant. 

So far, after importing this dataset, I cleaned it up by keeping only the variables of interest, and making the entry variable logical (having true/false values), as opposed to character (having yes/no values). 

The current dimensions of the data set are 1868 rows x 19 columns, and I do not believe that these data are tidy. 

## Answering some questions about the data 

```{r}
transit_df |> distinct(line, station_name) 
```

Using the distinct function, I can tell that there are 465 distinct stations. This code removed any duplicate rows so it shows only the number of distinct stations that are identified by name and line in the dataset. 

```{r}
transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name, ada)
```

By filtering out the distinct data on how many stations there are for only the ones where ADA compliance is "TRUE" I can tell there are 84 stations that are ADA complaint. 

```{r}
transit_df |>
  mutate(
    vending = case_match(
      vending, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE 
    )
  ) |>
  filter(vending == FALSE) 
```

First filtering out how many stations do not have vending

```{r}
transit_df |>
  mutate(
    vending = case_match(
      vending, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE 
    )
  ) |>
  filter(vending == FALSE, entry == TRUE) 
```

Second filtering out how many stations do not have vending and allow entrance

By first filtering out how many entrances / exits do not have vending (183), and then filtering out how many entrances / exits do not have vending and allow entry (69), I can find the proportion of station entrances / exits without vending that allow entrance. This would be 69 / 183, which I can compute in the console, to get 0.3770492 or 37.71% of station entrances / exits without vending allow entrance. 

## Reformatting the data 

Reformatting the data so that route number and route name are distinct variables 

```{r}
transit_tidy_df = 
  transit_df |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
  ) |>
  drop_na(route_name) 

view(transit_tidy_df)
```

## Answering more questions

```{r}
transit_tidy_df |>
  filter(route_name == "A")|>
  distinct(line, station_name)
```

By removing any duplicate rows so it shows only the number of distinct stations that are identified by name and line in the dataset, and filtering for the stations that serve only the A train, I can see there are 60  distinct stations serve the A train. 

```{r}
transit_tidy_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name, ada)
```

By removing any duplicate rows so it shows only the number of distinct stations that are identified by name and line in the dataset, and filtering for the stations that serve only the A train, and then from those filtering for the ones where ada compliance is true, I can see that of the stations that serve the A train, 17 are ADA compliant. 


