---
title: "P8105 HW2 - Sophie Gakis (sig2127)"
output: github_document
date: "2026-10-01"
---

# Problem 1 

Loading the tidyverse to start

```{r}
library(tidyverse)
```

## Importing the NYC Transit dataset, reading, and cleaning the data.  

```{r}
transit_df = 
  read_csv(
    file = "Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
    janitor::clean_names() |>
    select(line, station_name, station_latitude, station_longitude,  starts_with("route"), entry, vending, entrance_type, ada) |>
    mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE 
    )
  )
```

## Writing a paragraph to describe the data 

The `transit_df` dataset now contains `r ncol(transit_df)` columns. It contains subway lines, station names, latitude/longitude values for each station, what routes are served by each station, what entrace type is in the station, if there is vending and entry at the station, and if the station is ADA compliant. 

So far, after importing this dataset, I cleaned it up by keeping only the variables of interest, and making the entry variable logical (having true/false values), as opposed to character (having yes/no values). 

The current dimensions of the data set are `r nrow(transit_df)` rows x `r ncol(transit_df)` columns, and I do not believe that these data are tidy. 

## Answering some questions about the data 

```{r}
transit_df |> distinct(line, station_name) 
```

Using the distinct function, I can tell that there are 465 distinct stations. This code removed any duplicate rows so it shows only the number of distinct stations that are identified by name and line in the dataset. 

```{r}
transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name, ada)
```

By filtering out the distinct data on how many stations there are for only the ones where ADA compliance is "TRUE" I can tell there are 84 stations that are ADA complaint. 

```{r}
transit_df |>
  mutate(
    vending = case_match(
      vending, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE 
    )
  ) |>
  filter(vending == FALSE) 
```

First filtering out how many stations do not have vending

```{r}
transit_df |>
  mutate(
    vending = case_match(
      vending, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE 
    )
  ) |>
  filter(vending == FALSE, entry == TRUE) 
```

Second filtering out how many stations do not have vending and allow entrance

By first filtering out how many entrances / exits do not have vending (183), and then filtering out how many entrances / exits do not have vending and allow entry (69), I can find the proportion of station entrances / exits without vending that allow entrance. This would be 69 / 183, which I can compute in the console, to get 0.3770492 or 37.71% of station entrances / exits without vending allow entrance. 

## Reformatting the data 

Reformatting the data so that route number and route name are distinct variables 

```{r}
transit_tidy_df = 
  transit_df |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
  ) |>
  drop_na(route_name) 
```

## Answering more questions

```{r}
transit_tidy_df |>
  filter(route_name == "A")|>
  distinct(line, station_name)
```

By removing any duplicate rows so it shows only the number of distinct stations that are identified by name and line in the dataset, and filtering for the stations that serve only the A train, I can see there are 60  distinct stations serve the A train. 

```{r}
transit_tidy_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name, ada)
```

By removing any duplicate rows so it shows only the number of distinct stations that are identified by name and line in the dataset, and filtering for the stations that serve only the A train, and then from those filtering for the ones where ada compliance is true, I can see that of the stations that serve the A train, 17 are ADA compliant. 

# Problem 2 

## Reading in and cleaning the Mr. Trash Wheel dataset

```{r}
library(readxl)
```

```{r}
mr_trash_df = 
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
    sheet = "Mr. Trash Wheel", 
      range = "A2:N742",               
        na = c("", "NA", ".")
          ) |>
  janitor::clean_names() |>
  mutate(
    sports_balls = round(sports_balls), 
    sports_balls = as.integer(sports_balls), 
    plastic_bottles = as.numeric(plastic_bottles), 
    year = as.numeric(year),
    set_name = "mr_trash"
  )
```

## Reading in and cleaning the Professor Trash Wheel and Gwynnda datasets similarly 

```{r}
professor_trash_df = 
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
    sheet = "Professor Trash Wheel", 
      range = "A2:M139",               
        na = c("", "NA", ".")
          ) |>
  janitor::clean_names() |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles), 
    set_name = "professor_trash" 
  )

gwynnda_trash_df = 
  read_excel("Data/202610 Trash Wheel Collection Data.xlsx", 
    sheet = "Gwynnda the Good Wheel of the W", 
      range = "A2:L394",               
        na = c("", "NA", ".")
          ) |>
  janitor::clean_names() |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles),
    set_name = "gwynnda_trash" 
  )
```

## Combining all three datasets by binding the rows 

```{r}
trash_combined_df = 
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_trash_df)
```


## Writing a paragraph about this data 

After combining these three datasets, the `trash_combined_df` dataset contains `r nrow(trash_combined_df)` observations and `r ncol(trash_combined_df)` variables. Some important variables within `trash_combined_df` are `dumpster` which gives an ID number for the dumpster being investigated, `month`, `year`, and `date` to identify when the measurements were made, `weight_tons` to describe the weight of the trash from the dumpster measured in tons, and things like `glass_bottles`, `plastic_bags`, and `wrappers` to identify how many of certain objects were found within the dumpster.

## Answering some questions about this data 

```{r}
professor_weight_total = 
  professor_trash_df |>
    pull(weight_tons) |>
    sum(na.rm = TRUE)
```

The total weight of trash collected by Professor Trash Wheel was `r round(professor_weight_total, 2)` tons.

```{r}
gwynnda_cigarette_total = 
  gwynnda_trash_df |>
    filter (month == "June", year == 2022) |>
    pull(cigarette_butts) |>
    sum(na.rm = TRUE)
```

The total number of cigarette butts collected by Gwynnda in June 2022 was `r (gwynnda_cigarette_total)`. 

# Problem 3 

## Importing, cleaning, and tidying the dataset of baseline demographics 

```{r}
baseline_df = 
  read_csv(
    file = "Data/MCI_baseline.csv", 
    na = c(".", "NA", "")) |>
    janitor::clean_names() |> 
    slice (-1) |>
    rename(
     Current_Age = age_at_the_study_baseline, 
     Study_ID = x1 , 
     Sex = x1_male_0_female, 
     Education = years_of_education,  
     Apoe4 = x1_apoe4_carrier_0_apoe4_non_carrier, 
     Age_At_Onset = age_at_the_onset_of_mci_missing_if_a_subject_remains_mci_free_during_the_follow_up_period
      ) |>
  mutate(
    Sex = case_match(
      Sex,  
      "1" ~ "Male", 
      "0" ~ "Female"
    ),
    Apoe4 = case_match(
      Apoe4, 
      "1" ~ "Apoe4 Carrier", 
      "0" ~ "Apoe4 Non-Carrier"
    ),
    Current_Age = as.numeric(Current_Age), 
    Age_At_Onset = as.numeric(Age_At_Onset)
  ) |> 
  filter(is.na(Age_At_Onset) | Age_At_Onset > Current_Age)
```

## Answer some questions about this dataset 

As far as the import process went, I first started by importing the CSV file from my data folder, telling R what to recognize as NA values, and cleaning up the given names. Then I looked at the dataset and removed the first row since it was headings and not data. Then I renamed the columns/variables to simpler titles that made more sense to me. I then used `case_match` ensure that sex and APOE4 carrier status were appropriately encoded as words and not as numbers. Finally, I filtered out the dataset for only rows that had `age_at_onset` equal to NA or greater than their current age as to remove any participants who do not meet the stated inclusion criteria (having no MCI at baseline). 

A relevant feature of the dataset in my opinion is if the participant is an Apoe4 carrier or not.  

There were `r nrow(baseline_df)` participants recruited, and of these, `r nrow(filter(baseline_df, Age_At_Onset > 0))` develop MCI. The average baseline age is `r round(mean(baseline_df$Current_Age, na.rm = TRUE), 1)`. The proportion of women in the study who are APOE4 carriers is `r nrow(filter(baseline_df, Sex == "Female" & Apoe4 == "Apoe4 Carrier")) / nrow(filter(baseline_df, Sex == "Female"))`. 


## Importing, cleaning, and tidying the dataset of longitudinally observed biomarker values

```{r}
amyloid_df = 
  read_csv(
    file = "Data/mci_amyloid.csv", 
    na = c(".", "NA", "")) |>
    janitor::clean_names() |> 
    slice (-1) |>
    rename(
     Study_ID = study_id, 
     Baseline = time_in_years_elapsed_since_the_study_baseline_to_the_visit_where_biomarker_amyloid_42_40_ratio_was_measured, 
     Time_2 = na_3,  
     Time_4 = na_4, 
     Time_6 = na_5, 
     Time_8 = na_6
      ) |>
  pivot_longer(
    Baseline:Time_8,
    names_to = "Time of Visit", 
    values_to = "Years Elapsed")
```

As far as the import process went, I first started by importing the CSV file from my data folder, telling R what to recognize as NA values, and cleaning up the given names. Then I looked at the dataset and removed the first row since it was headings and not data. Then I renamed the columns/variables to simpler titles that made more sense to me. Finally, I used `pivot_longer` to fix that the years elapsed between baseline and the visit where the biomarker ratio was measures was spread across five columns, which corresponded to five observation times.

Some features of this dataset that I find interesting are all of the measures of the time in years that has elapsed from the study baseline to the visit where the biomarker ratio was measured. There are 5 measurements for this, baseline, time 2, time 4, time 6, and time 8. 

## Checking whether some participants appear in only the baseline or amyloid dataset

```{r}
filter(baseline_df, !Study_ID %in% amyloid_df$Study_ID)

filter(amyloid_df, !Study_ID %in% baseline_df$Study_ID)
```

The first filtration is looking for study ID's that only appear in the baseline dataset and not in the amyloid dataset. It looks like there are 8 of these. The second filtration is looking for study ID's that only appear in the amyloid dataset and not in the baseline dataset. It looks like there are 80 of these. 

## Combining both datasets so that only participants who appear in both datasets are retained

```{r}
combo_df = inner_join(baseline_df, amyloid_df, by = "Study_ID")
```

After combining these two datasets, the `combo_df` dataset contains `r nrow(combo_df)` observations and `r ncol(combo_df)` variables. This means that `r n_distinct(combo_df$Study_ID)` participants had data collected that appeared in both data sets. 

## Exporting the result as a CSV to my data directory

```{r}
write_csv(combo_df, "data/combo_baseline_and_amyloid.csv")
```

