---
title: "P8105 HW2 - Sophie Gakis (sig2127)"
output: github_document
date: "2026-10-01"
---

# Problem 1 

Loading the tidyverse to start

```{r}
library(tidyverse)
```

Importing the NYC Transit dataset, reading, and cleaning the data.  

```{r}
transit_df = 
  read_csv(
    file = "Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
    janitor::clean_names() |>
    select(line, station_name, station_latitude, station_longitude,  starts_with("route"), entry, vending, entrance_type, ada) |>
    mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE, 
      "NO" ~ FALSE 
    )
  )

nrow(transit_df)
ncol(transit_df)

view(transit_df)
```

The transit_df dataset now contains 19 variables (or columns). It contains subway lines, station names, latitude/longitude values for each station, what routes are served by each station, what entrace type is in the station, if there is vending and entry at the station, and if the station is ADA compliant. 

So far, after importing this dataset, I cleaned it up by keeping only the variables of interest, and making the entry variable logical (having true/false values), as opposed to character (having yes/no values). 

The current dimensions of the data set are 1868 rows x 19 columns, and I do not believe that these data are tidy. 

